Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-09-26 paper/main.tex
Shortening examples

\begin{lstlisting}[language=haskell,caption={\textbf{Haskell Mutation Tool}},label=lst:mutationTool]
{-# LANGUAGE FlexibleContexts #-}

module MutationTool where

import Language.Haskell.Exts

-- Main function
mutateAndTest :: String -> IO ()
mutateAndTest filePath = do
    parsedModule <- parseFile filePath
    case parsedModule of
        ParseFailed srcLoc err -> print err
        ParseOk modul -> do
            let mutatedModule = mutateModule modul
            writeFile "Mutated.hs" $ prettyPrint mutatedModule
            -- Here, you'd typically run your tests on "Mutated.hs"
            -- For now, just printing the mutated code:
            putStrLn $ prettyPrint mutatedModule

-- Perform mutation on a module
mutateModule :: Module -> Module
mutateModule = everywhere (mkT mutateExp)

-- Perform mutation on an expression
mutateExp :: Exp -> Exp
mutateExp (If cond trueBranch falseBranch) = If cond falseBranch trueBranch
mutateExp x = x

main :: IO ()
main = mutateAndTest "YourFile.hs"
\end{lstlisting}

\begin{lstlisting}[caption={\textbf{GPT-4 Comby Haskell Rules}},label=lst:haskellcomby]
head :[list] ==> last :[list]
tail :[list] ==> init :[list]
(:[expr1]) && (:[expr2]) ==> (:[expr1]) || (:[expr2])
(:[expr1]) || (:[expr2]) ==> (:[expr1]) && (:[expr2])
(:[expr1]) >>= (:[expr2]) ==> (:[expr1]) >> (:[expr2])
...
\end{lstlisting}

\begin{lstlisting}[caption={\textbf{GPT-4 Mutants on a simple Haskell Program}},label=lst:haskellmutants]
[0..] ==> [1..]
IO () ==> IO String  
"Current TODO list:" ==> "Updated TODO list:"
"No TODO entry matches the given number" ==> "No matching TODO found"
"Invalid command: " ==> "Unrecognized command: "
...
\end{lstlisting}

To understand how our approach compares with state of the art large language models (LLM), we prompt GPT-4 to generate a mutation testing tool for Haskell.
Surprisingly, the generated program in \cref{lst:mutationTool} has no syntax errors, and even runs to produce a mutant. However, the program only implements 
the inverting conditionals operation, a small subset of all the operators our tool implements. Furthermore, like existing mutation testing tools for source code,
the generated code performs AST manipulation in Haskell, making it harder to read than the comby rules of our approach. Asking the LLM to expand the set of rules,
 produces more operators, however the number of mutants generated by such an approach would be a small subset of what our approach is capable of doing.

>>>>

\begin{lstlisting}[language=haskell,caption={\textbf{GPT-4 Generated Haskell Mutation Tool}},label=lst:mutationTool]
-- Perform mutation on a module
mutateModule :: Module -> Module

-- Perform mutation on an expression
mutateExp :: Exp -> Exp
mutateExp (If cond trueBranch falseBranch) = If cond falseBranch trueBranch
mutateExp x = x
...
\end{lstlisting}

\begin{lstlisting}[caption={\textbf{GPT-4 Comby Haskell Rules}},label=lst:haskellcomby]
head :[list] ==> last :[list]
tail :[list] ==> init :[list]
(:[expr1]) && (:[expr2]) ==> (:[expr1]) || (:[expr2])
(:[expr1]) >>= (:[expr2]) ==> (:[expr1]) >> (:[expr2])
\end{lstlisting}

\begin{lstlisting}[caption={\textbf{GPT-4 Mutants on a simple Haskell Program}},label=lst:haskellmutants]
[0..] ==> [1..]
IO () ==> IO String  
"Current TODO list:" ==> "Updated TODO list:"
"Invalid command: " ==> "Unrecognized command: "
\end{lstlisting}

To understand how our approach compares with state of the art large language models (LLM), we prompt GPT-4 to generate a mutation testing tool for Haskell.
Surprisingly, the generated program in \cref{lst:mutationTool} has no syntax errors, and even runs to produce a mutant. However, the program only implements 
the inverting conditionals operation and some basic module mutations, a small subset of all the operators our tool implements. Furthermore, like existing mutation testing tools for source code,
the generated code performs AST manipulation in Haskell, making it harder to read than the comby rules of our approach. Asking the LLM to expand the set of rules,
 produces more operators, however the number of mutants generated by such an approach would be a small subset of what our approach is capable of doing.
